# Examining data I

## Introduction
Welcome to your second week of **Introduction to Quantitative Research Methods**. This week we will focus on **examining data** using measures of central tendency and measures of dispersion. These measures are collectively known as **descriptive statistics**. We will also talk about some basic data visualisation. Also, as by now everyone should be up and running with R, we will apply some of these descriptive measures onto some data. [Nice](https://www.youtube.com/watch?v=LBduNcf1eQc)!

This week is structured by **3** short videos, practical material that you need to work through in preparation for Thursday's seminar, and a seminar task that you need to do in preparation for Thursday's seminar.

Let's get to it.

#### Video: Introduction W02 {-}
```{r 02-short-lecture-welcome, warnings=FALSE, message=FALSE, echo=FALSE}
library(vembedr)
embed_msstream('40cef170-a483-44bf-b7ca-5b08ca4a790e') %>% use_align('left')
```
[Lecture slides] [[Watch on MS stream]](https://web.microsoftstream.com/video/40cef170-a483-44bf-b7ca-5b08ca4a790e)

```{r 02-settings, warnings=FALSE, message=FALSE, echo=FALSE}
# settings
options(max.print=30)
```

### Reading list {#reading-list-w02}
Please find the reading list for this week below. We strongly recommend that you read the **core reading material** before you continue with the rest of this week's material.

#### Core reading {-}
- Lane *et al.*, 2003, **Chapter 3: Summarizing Distributions**. In: Lane *et al.*, 2003, Introduction to Statistics. Houston, Texas:  Rice University. [[Link]](http://onlinestatbook.com/Online_Statistics_Education.pdf)

#### Supplementary reading {-}
- Diez *et al.*, 2019, **Chapter 2: Summarizing data**. In: Diez *et al.*, 2019, OpenIntro Statistics. Fourth Edition. Boston, Massachusetts: OpenIntro.
[[Link]](https://leanpub.com/openintro-statistics)

### Q&A session
This week there is **NO** live Q&A session scheduled. Please post any questions you have on the [POLS0008 Forum](https://moodle.ucl.ac.uk/mod/forum/view.php?id=1632137) or ask them to one of the tutors during your live seminar on Thursday.

## Measures of central tendency
Any research project involving [quantitative data](https://en.wikipedia.org/wiki/Quantitative_research) should start with an exploration / examination of the available data sets. This applies both to data that you have collected yourself and data that you have acquired in a different way, e.g. through downloading official [UK Census and labour market statistics](https://www.nomisweb.co.uk/). The set of techniques that is used to examine your data in first instance is called [descriptive statistics](https://en.wikipedia.org/wiki/Descriptive_statistics). Descriptive statistics are used to describe the basic features of your data set and provide simple summaries about your data. Together with simple visual analysis, they form the basis of virtually **every** quantitative data analysis.

#### Video: Measures of central tendency {-}
```{r 02-descriptive-statistics, warnings=FALSE, message=FALSE, echo=FALSE}
embed_msstream('a6169cec-43ad-4c3f-bda6-9cc8fc4575f5') %>% use_align('left')
```
[[Lecture slides]](https://liveuclac-my.sharepoint.com/:b:/g/personal/ucfajtv_ucl_ac_uk/EdQtMgoDs4BPkvVXKqXPDjcBP1OllLAePoqYQM0JO0I8Yw?e=YvMuCS) [[Watch on MS stream]](https://web.microsoftstream.com/video/a6169cec-43ad-4c3f-bda6-9cc8fc4575f5)

For this tutorial we will initially continue to use the `London.Pop` object that we created during last week's tutorial. You may still have it loaded into your R workspace if your RSTudio Server session is still active. To check if you do, you can use the `ls()` command to get a list of all objects that are available in your RStudio session. Type this into the command line and see if `London.Pop` is printed. 

If not you can simply reload it:
```{r 02-make-sure-working-dir-is-set, warnings=FALSE, message=FALSE, eval=FALSE}
# set the working directory path to the folder you wish to use
# this would be the folder that you created during last week's material
setwd('~/POLS0008') 
# note the single / (\\ will also work)
```

If you struggle with setting up your working directory and you need a reminder, [have a look at how we did this last week](understanding-data.html#working-directory)!

```{r 02-load-csv, warnings=FALSE, message=FALSE, eval=FALSE}
# load csv file from working directory
London.Pop <- read.csv('census-historic-population-borough.csv')
```

```{r 02-load-csv-for-realz, warnings=FALSE, message=FALSE, echo=FALSE}
# load csv file from working directory
London.Pop <- read.csv('raw/census-historic-population-borough.csv')
```

Use the `head()`, or `View()`, function to remind yourself of the structure of the London population data frame. You should see **25** columns of data.

#### Questions {-}
1. What is the difference between using `head()` and `View()` to inspect your data?
2. Can you combine the functions `View()` and `head()` to browse only through a few lines of your data?

The `mean()` and `median()` were some of the first R functions we used at the last week to describe our `Friends` data set. Now we know a little more about these **measures of central tendency**, we can also apply them on our `London.Pop` data set. Let's do this by focusing on London's population in 2011.

```{r 02-slice-the-population, warnings=FALSE, message=FALSE}
# select the relevant columns
London.Pop2011 <- London.Pop[,c(1:2,24)]
```

#### Questions {-}
1. Do you remember what the `c()` function does? Why do we need to use the `c()` in this case?
2. Instead of using the `[ , ]` [syntax](https://en.wikipedia.org/wiki/Syntax) to slice our data frame, how can we create the same `London.Pop2011` object by using the `$` syntax?

Now we can calculate our **measures of central tendency** using R's built-in functions for the median and the mean.

```{r 02-calculate-measures, warnings=FALSE, message=FALSE }
# calculate the median of the 2011 population variable
median(London.Pop2011$Persons.2011)

# calculate the mean of the 2011 population variable
mean(London.Pop2011$Persons.2011)
```

#### Questions {-}
1. How do you explain that the median is larger than the mean for this variable?

R does not have a standard in-built function to calculate the mode. As we still want to show the mode, we create a user function to calculate the mode of our data set. This function takes [a numeric vector](https://r4ds.had.co.nz/vectors.html) as input and gives the mode value as output.

<div class="note">
You do not have to worry about creating your own functions, so just copy and paste the code below to create the `get_mode()` function. 
</div>

```{r 02-create-a-mode-function, warnings=FALSE, message=FALSE }
# create a function to calculate the mode
get_mode <- function(x) {
  
  # get unique values of the input vector
  uniqv <- unique(x)
  
  # select the values with the highest number of occurences
  uniqv[which.max(tabulate(match(x, uniqv)))]
}

# calculate the mode of the 2011 population variable
get_mode(London.Pop2011$Persons.2011)
```

#### Questions {-}
1. What is the level of measurement of our *Persons.2011* variable? Nominal, ordinal, or interval/ratio?
2. Even though we went through all the trouble to create our own function to calculate the mode, do you think it is a good choice to calculate the mode for this variable? Why? Why not?

Although R does most of the hard work for us, especially with the `mean()` and the `median()` function, it is a good idea to once go through the calculations of these two central tendency measures ourselves. Let's start by step-by-step calculating the mean and then verify our results with the results of R's `mean()` function.

```{r 02-verify-that-mean, warnings=FALSE, message=FALSE }
# get the sum of all values
Persons.2011.Sum <- sum(London.Pop2011$Persons.2011)

# inspect the result
Persons.2011.Sum

# get the total number of observations
Persons.2011.Obs <- length(London.Pop2011$Persons.2011)

# inspect the result 
Persons.2011.Obs

# calculate the mean
Persons.2011.Mean <- Persons.2011.Sum / Persons.2011.Obs

# inspect the result
Persons.2011.Mean

# compare our result with R's built-in function
mean(London.Pop2011$Persons.2011) == Persons.2011.Mean
```

Great. Our own calculation of the mean is identical to R's built-in function. Now let's do the same for the median.

```{r 02-verify-that-median, warnings=FALSE, message=FALSE }
# get the total number of observations
Persons.2011.Obs <- length(London.Pop2011$Persons.2011)

# inspect the result
Persons.2011.Obs

# order our data from lowest to highest
Persons.2011.Ordered <- sort(London.Pop2011$Persons.2011, decreasing=FALSE)

# inspect the result
Persons.2011.Ordered

# get the number of the observation that contains the median value
Persons.Median.Obs <- (Persons.2011.Obs + 1)/2

# inspect te result
Persons.Median.Obs

# get the median
Persons.2011.Median <- Persons.2011.Ordered[Persons.Median.Obs]

# inspect the result
Persons.2011.Median

# compare our result with R's built-in function
median(London.Pop2011$Persons.2011) == Persons.2011.Median
```

#### Questions {-}
1. For the calculation of the median: what does the `decreasing=FALSE` parameter in the `sort()` function do? What happens if we change this to `decreasing=TRUE`? Does this affect the results for the calculation of the median? Why? Why not?
2. When extracting the median we using `[Persons.Median.Obs]` and not `[17]` to select our value. Although the results are identical, what would be a good reason to select the first option over the second option? 
3. When comparing our own calculation to R's built-in function we are using two 'equal to' signs `==`, what happens if we only use one 'equal to' `=` sign?

## Simple plots
Before moving on to the second set of descriptive statistics, the **measures of dispersion**, simple data visualisations are also an extremely powerful tool to getting to know your data. In fact, tools to create high quality plots have become one of R's greatest assets. This is a relatively recent development since the software has traditionally been [focused on the statistics](https://en.wikipedia.org/wiki/R_(programming_language)) rather than visualisation. The standard installation of R has base graphic functionality built in to produce very simple plots. For example we can plot the relationship between the London population in 1811 and 1911. 

<div class="note">
Next week we will be diving deeper into data visualisation and making plots and graphs in R, but for now it is a good idea to already take a sneak peek at how to create some basic plots.
</div>

```{r 02-quick-plot, warnings=FALSE, message=FALSE}
# make a quick plot of two variables of the London population data set
plot(London.Pop$Persons.1811,London.Pop$Persons.1911)
```

#### Questions {-}
1. What happens if you change the order of the variables you put in the `plot()` function? Why?
2. Instead of using the `$` to select the columns of our data set, how else can we get the same results?

The result of calling the `plot()` function, is a very simple [scatter graph](https://en.wikipedia.org/wiki/Scatter_plot). The `plot()` function offers a huge number of options for customisation. You can see them using the `?plot` help pages and also the `?par` help pages (`par` in this case is short for parameters). There are some examples below (note how the parameters come after specifying the `x` and `y` columns).

```{r 02-quick-plot-with-some-options, warnings=FALSE, message=FALSE}
# add a title, change point colour, change point size
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main='Quick Plot in R', col='blue', cex=2)
```

```{r 02-quick-plot-with-different-options, warnings=FALSE, message=FALSE}
# add a title, change point colour, change point symbol
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main="Another Quick Plot in R", col='magenta', pch=22)
```

<div class="note">
For more information on the plot parameters (some have obscure names) have a look here: http://www.statmethods.net/advgraphs/parameters.html
</div>

#### Recap {-}
In this section you have learnt how to:

1. Calculate the mode, median, and mean of in R.
2. Make some simple scatter plots (in preparation for next week) to visualise your data.

## Measures of dispersion
When getting to know your data, **measures of central tendency** alone are not enough as they only tell you what a 'typical' value looks like but they do not tell you anything about all other values. Therefore we also need to look at some **measures of dispersion**. Measures of dispersion describe the spread of data around a central value (e.g. the mean, the median, or the mode). The most commonly used measure of dispersion is the [standard deviation](https://en.wikipedia.org/wiki/Standard_deviation). The standard deviation is a measure to summarise the spread of your data around the mean. The short video below will introduce you to the standard deviation as well as to three other measures of dispersion: the **range**, the **interquartile range**, and the **variance**.

#### Video: Measures of dispersion {-}
```{r 02-visualising-data, warnings=FALSE, message=FALSE, echo=FALSE}
embed_msstream('8623cbd8-ab11-40f9-b987-f7e966db50db') %>% use_align('left')
```
[[Lecture slides]](https://liveuclac-my.sharepoint.com/:b:/g/personal/ucfajtv_ucl_ac_uk/EQBtLckmEItBniy4Jh_toSEBJkOMZ5BKl5pid5BBvRNz0A?e=B01n4f) [[Watch on MS stream]](https://web.microsoftstream.com/video/8623cbd8-ab11-40f9-b987-f7e966db50db)

For the rest of this tutorial we will change our data set to one containing the number of assault incidents that ambulances have been called to in London between 2009 and 2011. You will need to download a prepared version of this file called: `ambulance_assault.csv` and upload it to your working directory. It is in the same data format (`csv`) as our London population file so we use the `read.csv()` command again.

#### File download {-}
| File                                                 | Type           | Link |
| :------                                              | :------        | :------ |
| Assault Incidents London        | `csv`             | [Download](https://liveuclac-my.sharepoint.com/:x:/g/personal/ucfajtv_ucl_ac_uk/EcWgTqy0uABHg_JeJLKgTS4BsRfGI3wcgjfuId_srvBtlg?e=NywiNE) |

```{r 02-load-ambulance-csv, warnings=FALSE, message=FALSE, eval=FALSE}
# load csv file from working directory
input <- read.csv('ambulance-assault.csv')
```

```{r 02-load-ambulance-csv-for-realz, warnings=FALSE, message=FALSE, echo=FALSE}
# load csv file from working directory
input <- read.csv('raw/ambulance-assault.csv')
```

```{r 02-check-ambulance-csv-for-realz, warnings=FALSE, message=FALSE}
# inspect the results 
head(input)
```

```{r 02-check-nrow-ambulance-csv-for-realz, warnings=FALSE, message=FALSE}
# inspect the size of the data set
nrow(input)
```

You will notice that the data table has 4 columns and 649 rows. The column headings are abbreviations of the following:

| Column heading | Full name | Description |
| :---           | :---      | :---------- |
|**BorCode** | Borough Code | London has 32 Boroughs (such as Camden, Islington, Westminster, etc.) plus the City of London at the centre. These codes are used as a quick way of referring to them from official data sources. 
| **WardName** | Ward Name | Boroughs can be broken into much smaller areas known as [Wards](https://en.wikipedia.org/wiki/Wards_and_electoral_divisions_of_the_United_Kingdom). These are electoral districts and have existed in London for centuries. | 
| **WardCode** | Ward Code | A statistical code for the Wards above. |
| **WardType** | Ward Type | A classification that groups wards based on similar characteristics. |
| **Assault_09_11** | Assault Incidents | The number of assault incidents requiring an ambulance between 2009 and 2011 for each Ward in London. |

Let's start by calculating two measures of central tendency by using the `median()` and `mean()` functions.

```{r 02-descriptive-statistics-ambulance, warnings=FALSE, message=FALSE}
# calculate the median of the assault incident variable
median(input$Assault_09_11)

# calculate the mean of the assault incident variable
mean(input$Assault_09_11)
```

#### Questions {-}
1. How do you explain that the mean is larger than the median for this variable?

Great. Let's now calculate some measures of dispersion for our data: the range, the interquartile range, and the standard deviation. The calculation of the range is very straightforward as we only need to subtract the the minimum value from the the maximum value. We can find these values by using the built-in `min()` and `max()` functions.

```{r 02-calculate-that-range, warnings=FALSE, message=FALSE}
# get the minimum value of our data
min(input$Assault_09_11)

# get the maximum value of our data
max(input$Assault_09_11)

# calculate the range
1582 - 0

# or in one go
max(input$Assault_09_11) - min(input$Assault_09_11)
```

#### Questions {-}
1. Why do we get a value of `0` when executing our `min()` function?
2. What does this range mean? 

The interquartile range requires a little bit more work as we now need to work out the values of the 25th and 75th [percentile](https://en.wikipedia.org/wiki/Percentile).

<div class="note">
A percentile is a score at or below which a given percentage fall. For example, the 50th percentile (also known as the median!) is the score at or below which 50% of the scores in the distribution may be found.
</div>

```{r 02-calculate-that-interquartile-range, warnings=FALSE, message=FALSE}
# get the total number of observations
Input.Obs <- length(input$Assault_09_11)

# inspect the result
Input.Obs

# order our data from lowest to highest
Input.Ordered <- sort(input$Assault_09_11, decreasing=FALSE)

# inspect the result
Input.Ordered

# get the number ('index value') of the observation that contains the 25th percentile
Input.Q1 <- (Input.Obs + 1)/4

# inspect the result
Input.Q1

# get the number ('index value') of the observation that contains the 75th percentile
Input.Q3 <- 3*(Input.Obs + 1)/4

# inspect the result
Input.Q3

# get the 25th percentile
Input.Ordered[163]

# get the 75the percentile
Input.Ordered[488]

# get the interquartile range
233 - 86
```

As shown in the short video, we can also visually represent our range, median, and interquartile range using a box and whisker plot:

```{r 02-calculate-that-interquartile-range, warnings=FALSE, message=FALSE}
# make a quick boxplot of our assualt incident variable
boxplot(input$Assault_09_11, horizontal=TRUE)
```

#### Questions {-}
1. There is a big difference between the range that we calculated and the interquartile range that we calculated. What does this mean?
2. The 25th and 75th percentile do not return integer but a fraction. Why do we use 163 and 488 to extract our percentile values and not 162 and 487?






These are commonly used descriptive statistics. To make things even easier, R has a `summary()` function that calculates a number of these routine statistics simultaneously.

```{r 02-descriptive-statistics-summary, warnings=FALSE, message=FALSE}
# calculate the most common descriptive statistics for the assault incident variable
summary(input$Assault_09_11)
```




You should see you get the minimum (Min.) and maximum (Max.) values of the *assault_09_11* column; its first (1st Qu.) and third (3rd Qu.) quartiles that comprise the interquartile range; its the mean and the median. The built-in R `summary()` function does not calculate the standard deviation. There are functions in other libraries that calculate more detailed descriptive statistics, including `describe()` in the `psych` package, which we will use in the later tutorials. We can also use the `summary()` function to describe a categorical variable and it will list its levels:

```{r 02-descriptive-statistics-categorical-variable, warnings=FALSE, message=FALSE}
# summarise a categorical variable
summary(input$WardType)
```

## Seminar {#seminar-w02}

Please find the seminar task and seminar questions for this week's seminar below. 

<div class="note">
**Please make sure that you have executed the seminar task and have answered the seminar questions before the seminar!**
</div>

#### Seminar task {-}

#### Seminar questions {-}

1. Explain why each of these statistics are useful and what type of data are required to calculate them:
- Mode
- Median
- Mean
- Range
- Interquartile Range
- Standard Deviation

#### Seminar link {-}
Seminars for all groups take place on **Thursday morning**. You can find the **Zoom** link to your seminar group on [Moodle](https://moodle.ucl.ac.uk/course/view.php?id=6845). 

## Before you leave {#byl-w02}
Save your R script by pressing the **Save** button in the script window. [That is it for this week](https://www.youtube.com/watch?v=Wmc8bQoL-J0)!